Relationships in the extremes and their influence on competition and coexistence

### 2021-09-20 paper/Paper.Rnw
new para4: mechanics of extreme events and ATA

%***DAN: Jasmin to edit the below as follows:
%1) Topic sentence: Another potentially important reason to quatify the importance of ATA to species coexistence is that there may
%be reason to believe that climate change can alter ATAs of environmental variables implicated in coexistence, as follows. 
%2) Just adapt below text, maybe try to simplify a bit.
%3) This para could just focus on the mechanics of how ATAs change for two variables measured at different times. 
%***DAN: Keeping this sentence so Jasmin can put it somwhere, not here.
%Given that storage effect depends on the relationship between E and C, we question the consequences of EC relationships characterized by ATA, the presence of which may be increasing as climate change brings about more extreme climatic events. 
%***DAN: I am also having a hard time identifying the topic sentence of this paragraph, can you tell me which one it is, please,
%in a comment right under this one?
%**JAS: I intended for the first sentence to be the topic sentence 
% a better one:
% Extreme events can have important ecological impacts for storage effect and coexistence by introducing ATAs. 
Extreme events can have important ecological impacts for
competing communities where species growth and the maintenance of biodiversity depends on distinct temporal events/dynamics. If extreme events are more temporally extensive than moderate disturbances, then it is more likely that two species will be influenced by the same events. Both species may become more alike (correlated), an important consideration for competition and coexistence. For example, an extreme flood event that lasts across the growth season of two different species and increases recruitment in both species will cause recruitment of the two species to become more correlated in the upper tail end of the distribution. A subsequent change in competitive ability and coexistence is not just attributed to the extreme event alone, but from the resulting extreme relationship between species (ATA). 
%***DAN: Leave the following for the next para
Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in general (see above), they do not directly consider ATA between species response caused by environmental changes -- extreme relationships caused by extreme events (but see \cite{Ghosh_2021} for impact of ATA on community aggregate properties).

%***DAN:
%1) Topic sentence: To make the above ideas more concrete, we review a relevant example from the literature that may be
%influenced as described above. 
%2) Make it concrete - name the system, name the organisms, name the seasons. 
To contextualize idea of how ATA in environmental response might manifest and modfiy the outcome of coexistence, we introduce an example of a two species grass-forb plant community (inspired by \cite{Chesson_2012}). The community follows the criteria required for temporal storage effects to operate. The forb dominates and has high fecundity when a sufficient threhold of rain is reached in the early season. The grass dominates and has high fecundity when there is sufficeint rain in the late season. Competition is influenced by resource comsumption of growing juvenile plants. Dormant seeds are unaffected by competiton and can persist in the community in an unfavorable environment. When rain fall dynamics temporally fluctuate from year to year, relative abundance and dominance of the grass-forb community also fluctuate year to year. The result is a relationship between the grass's and forb's environmental responses (fecundity) that is symmetrically associated, and, following our simulated example (Figure \ref{fig:pedagog}), the species persist in coexistence. However, if environmental responses become asymmetrically associated by, say, extreme rainfall events, coexistence can be altered. If the community is disturbed by frequent and long floods with some years of normalcy, the grass-forb environmental response relationship may become more correlated in the higher values or right-tail associated. Conversely, if the community is disturbed by frequent and long drought with some years of normalcy, the grass-forb environmental response relationship may become more correlated in the lower values or left tail associated. Again, we see the possible consequences of ATA in Figure \ref{fig:pedagog}, i.e., less or no coexistence.

>>>>

%***DAN: Jasmin to edit the below as follows:
%1) Topic sentence: Another potentially important reason to quatify the importance of ATA to species coexistence is that there may
%be reason to believe that climate change can alter ATAs of environmental variables implicated in coexistence, as follows. 
%2) Just adapt below text, maybe try to simplify a bit.
%3) This para could just focus on the mechanics of how ATAs change for two variables measured at different times. 
%***DAN: Keeping this sentence so Jasmin can put it somwhere, not here.
%Given that storage effect depends on the relationship between E and C, we question the consequences of EC relationships characterized by ATA, the presence of which may be increasing as climate change brings about more extreme climatic events. 
%***DAN: I am also having a hard time identifying the topic sentence of this paragraph, can you tell me which one it is, please,
%in a comment right under this one?
% Extreme events can have important ecological impacts for storage effect and coexistence by introducing ATAs. 
Furthermore, an additional driving reason to quantify the significance of ATAs for species coexistence is the potential for climate change to alter ATAs of environmental variables implicated in coexistence, as follows. 
The consequence of climate change that has the most potential to influence ATAs in the environment the increase in intensity, duration, extent, and frequency of extreme climatic events, i.e., heat-waves, floods, hurricanes, fires, etc. 
Temporally extensive extreme events can have important ecological impacts for
competing populations where species growth and the maintenance of biodiversity is dependent of phenologically separated environmental signals. A temporally extensive event can equalize the phenologically separated environmental signals, making those phenological signals more correlated. For example, consider two precipitations signals, one in early spring and one in late summer. If a long flood occurs, it's likely that both signals will be close to equal in that year. If extreme floods occur in more years than not, then the two signal will become more correlated in higher precipitation values, a right-tailed ATA. Similarly, if extreme droughts occur in more years than not, then the signals will be more correlated in the lower values, a left-tailed ATA. The same can be said about spatially separated signals and more spatially extensive extreme events. ATA of environmental signals can then introduce or change ATAs in the fitness or traits of species that depend on the signals and potentially affect their coexistence.
%***DAN:
%1) Topic sentence: To make the above ideas more concrete, we review a relevant example from the literature that may be
%influenced as described above. 
%2) Make it concrete - name the system, name the organisms, name the seasons. 
To contextualize idea of how ATA in environmental response might manifest and modfiy the outcome of coexistence, we introduce an example of a two species grass-forb plant community (inspired by \cite{Chesson_2012}). The community follows the criteria required for temporal storage effects to operate. The forb dominates and has high fecundity when a sufficient threhold of rain is reached in the early season. The grass dominates and has high fecundity when there is sufficeint rain in the late season. Competition is influenced by resource comsumption of growing juvenile plants. Dormant seeds are unaffected by competiton and can persist in the community in an unfavorable environment. When rain fall dynamics temporally fluctuate from year to year, relative abundance and dominance of the grass-forb community also fluctuate year to year. The result is a relationship between the grass's and forb's environmental responses (fecundity) that is symmetrically associated, and, following our simulated example (Figure \ref{fig:pedagog}), the species persist in coexistence. However, if environmental responses become asymmetrically associated by, say, extreme rainfall events, coexistence can be altered. If the community is disturbed by frequent and long floods with some years of normalcy, the grass-forb environmental response relationship may become more correlated in the higher values or right-tail associated. Conversely, if the community is disturbed by frequent and long drought with some years of normalcy, the grass-forb environmental response relationship may become more correlated in the lower values or left tail associated. Again, we see the possible consequences of ATA in Figure \ref{fig:pedagog}, i.e., less or no coexistence.

are more temporally extensive than moderate disturbances, then it is more likely that two species will be influenced by the same events. Both species may become more alike (correlated), an important consideration for competition and coexistence. For example, an extreme flood event that lasts across the growth season of two different species and increases recruitment in both species will cause recruitment of the two species to become more correlated in the upper tail end of the distribution. A subsequent change in competitive ability and coexistence is not just attributed to the extreme event alone, but from the resulting extreme relationship between species (ATA). 
%***DAN: Leave the following for the next para
Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in general (see above), they do not directly consider ATA between species response caused by environmental changes -- extreme relationships caused by extreme events (but see \cite{Ghosh_2021} for impact of ATA on community aggregate properties).
